# (PART) Data analysis: A First Encounter {-}

### {-}

In this section we will tackle a typical data analysis problem: determining if two groups, such as organisms or drug treatments, an be considered statistically different from each other.  We will use data from a paper titled "Sperm competition and the evolution of precopulatory weapons: Increasing male density promotes sperm competition and reduces selection on arm strength in a chorusing frog" by Buzatto et al (2015).  

The end goal is to compare the size of the arms on female and male frogs.  First, though, we will get to know the data by calculating summary statistics and making exploratory graphs.  We will then carry out a t-test and grapple with with the meaning and interpretation of the output  Finally, we'll explore how best to plot the output of a t-test.

Section outline:

1. Data exploration with summary statistics
1. Graphical data exploration with boxplots
1. Plotting means and measures of variation and precision
1. T-tests
1. Plotting the output of a t-test


# Data Analysis Encounter: Summary Statistics


```{r , include=F, echo = F}
knitr::opts_chunk$set(echo = TRUE,
                      tidy = F, 
                      warning = F,
                      error = F,
                      message = F,
                      #cache = TRUE,
                      fig.margin = TRUE,
                      #results="hide",
                      fig.width = 4,
                      fig.height = 3)
```

| **Nathan Brouwer, Phd**
| brouwern at gmail.com
| https://github.com/brouwern
| Twitter: lobrowR

## Introduction

### Goals and objectives

Get to know the **frogarms** dataset and learn how to calculate summary statistics using basic R functions (eg **summary()** ) and also with the handy tools in the **dplyr**.

### Outline

1. Load packages and data
1. Subset unique dataset
1. Calculate summary stats on columns
1. Use dplyr
1. Calculate summary stats by groups using dplyr

### Packages

* devtools
* wildlifeR (from GitHub)
* dplyr

### Functions

* devtools::install_github
* wildlifeR::make_my_data2L
* dim, nrow, ncol
* head, tail
* names
* summary
* median
* min, max, range
* var, sd
* nrow, length (for sample size)
* dplyr::summarise, dplyr::summarize
* group_by
* dplyr::n

## Preliminaries

First, we need to install the necessary packages.  The data are in a package stored on GitHub called **wildlifeR**.  The **devtools** package is needed for downloading from Github.  We'll also use **dplyr** for grouping data and calculating summary statistics.

### Load packages

You might have to install or re-install wildlifeR using **install.packages()**.  **If you have done this recently you can skip this step.**

```{r, first.chnk.sxn4.ch1 , eval=F}
library(devtools)

install_github("brouwern/wildlifeR") #Note that text is quoted " "
```

Recall that *downloading* a package and actually *loading* it into R's active memory are different things.  To actually use the package you need to use the **library()** command to load it into memory.
```{r echo = F}
library(wildlifeR)
```

### Load data

The data we'll be using is in a dataset called "frogarms" in the wildlifeR package.

```{r}
data(frogarms) #[_]
```

You can find out information about these data using the ? command.  Note that there are no parentheses required for this ( "?(frogarms)" is wrong)

```{r echo = F}
?frogarms
```


### Subset your data

In this lesson we'll be primariy working with a personalized subset of the data.  This will allow us to 

1. See the effects of sample size by comparing the larger frogarms data to your subset
1. See the effects of random variation on things like p-values

The worksheet that accompanies this chapter is meant to facilitate these comparisons between the full (frogarms) and sub datasets, and also between you can classmates.  The code that follows is focused on working with the subset we will generate below, but the same commands should also be run on the full frogarms dataset.

The function **make_my_data2L()** in the wildlfieR will extract out a random subset of the data.  Change "my.code" to your school email address, minus the "@ pitt.edu" or whatever your affiliation is.
```{r}
my.frogs <- frogarms # [_]
my.frogs <- make_my_data2L(dat = frogarms,
                           my.code = "nlb24", # <=  change this!
                           cat.var = "sex",
                           n.sample = 20,
                           with.rep = FALSE)

```

n.sample is set to 20.  This is set up to extract 20 unique individuals of each sex (20 male, 20 female).  Check that you dataframe is 2*20 = 40 rows using the dim() command.

```{r}
dim(my.frogs) # [_]
```


------------

### An aside on R functions (Optional)  [O]

The following sections are optional.  The 2sst task is easy to wrap your brain around (looking at the code behind a function); the 2nd is more advanced (debugging a function).

#### OPTIONAL (easy): Looking at the code behind a function

** This section is optional, but easy for beginners**

Functions in R are typically written using R code.  To see the underlying code you can type the function name in the console but nothing else, then run the command

```{r echo = T, eval = F}
make_my_data2L
```

The formatting of the output in the console might look a bit goofy; you can adjusted the conlse size so it looks better.  This code is fairly long but is mostly fairly basic R commands and is a single function.  In contrast, many R functions call other functions within them..

You can't always see the underlying code for a function though.  Try to look at the t.test function

```{r echo = T, eval = F}
t.test
```

### OPTIONAL (intermediate): Function defaults

**This section is optional**

>"Statistics is the science of defaults" (6 November 2012, https://andrewgelman.com)

See what happens when you run this code; note that there is **no** "my.code = ..." bit in it (just to simplify things).
```{r echo = T, eval = F}
make_my_data2L(dat = frogarms, cat.var = "sex")
```

Why does this work?  Check out the helpfile using **?make_my_data2L**, or look at the raw code as we just did.  Notice that right after the function name is listed the following things

* dat
* my.code = "nlb24"
* cat.var
* n.sample = 20
* with.rep = FALSE

These are the **arguements** that **make_my_data2L()** takes.  When an arugment name is followed by an "= ..." that means a default has been set.  If you call the function and don't specify what you want for a specific arguement, R checks for pre-specified **defaults** and uses those as needed.  Note that the use of defaults can be problematic, since you might use a deafult you didn't intend to.   Most defaults are sensible, and essential things like dataframes that need to be supplied rarely have defaults.


### OPTIONAL (Advanced): Debugging a function

**This section is optional and not relevant for beginners**

When a function is not working, or you want to understand how it works, you can debug it.  First, tell R that the next time you run the function you want to debug it

```{r echo = T, eval = F}
debugonce(make_my_data2L)
```

Then run the function

```{r echo = T, eval = F }
make_my_data2L()
```

This will create a new tab in **debugger mode**.  Every time you press "enter" you will step through the code to the next full line of R code (note that a line of functional R code can span more than one line of code in a file or when rendered on a screen).  When you get to the end of the function you will exit the debugger and go back to normal R mode.

If you want to interact with the function while its debugging you can type directly in the console.  For example, trying running the ls() command after every few lines of code to see what happens.  You can call dim(), summary(), is() etc on anything you find.



**END OPTIONAL SECTION**

-----------

## Getting to know your personalized dataframe

Let's get to know your personalized subset of the data.

### Dataframe dimension

The following commands tell you the row x column dimension, number of rows, and number of columns.

```{r eval = F}
dim(my.frogs)  #[_]
nrow(my.frogs) #[_]
ncol(my.frogs) #[_]
```

----------

### Optional: Accessing elements of objects

**The following section is optional.**

The commands dim(), nrow and ncol all generate objects that display information on the dimension of a dataframe.  dim() produces and object that is a "vector" that is 1 row x 2 elements in size.  We can select these individual elements using square brackets "[...]"

The full dim() output
```{r}
dim(my.frogs)
```

The 1st element of the dim() output is accessed by appending "[1]" to the very end of the line of code

```{r}
dim(my.frogs)[1] #"[1]" goes outside the ")"
```

The 2nd element

```{r}
dim(my.frogs)[2]
```

Both elements: use "[1:2]" with a ":".

```{r}
dim(my.frogs)[1:2]
```


Another way to get both elements, using "c(...)"

```{r}
dim(my.frogs)[c(1,2)]
```


What happens when you execute the command below command?  Why?
```{r}
dim(my.frogs)[c(2,1)]
```


If you want to know one reason why R can drive you crazy, run this code, without the "c" before "(1,2)"

```{r eval = F}
dim(my.frogs)[(1,2)]
```

But then try this; same as above but with ":" instead of "," between the numbers.
```{r eval = F}
dim(my.frogs)[(1:2)]
```

Yes, R is that picky.

**End optional section**

----------

### Dataframe preview

The **head()** and **tail()** commands give us short previews of the dataframe.

**head()** gives the top few rows
```{r eval = F}
head(my.frogs) #Note: I've truncated the output that I've actually shown
```

```{r echo = F}
head(my.frogs)[-1,-ncol(my.frogs)]
```


The bottom few rows with **tail()**
```{r eval = F}
tail(my.frogs)
```

```{r echo = F}
tail(my.frogs)[-1,-ncol(my.frogs)]
```

**names()** just gives of the names of the columns.  It is the same as **colnames()**.
```{r}
names(my.frogs)
```


Again, if you want to know what the column names are, use the ? command
```{r eval = F}
?my.frogs
```


## Summary statistics

**This section is review.  If you are familar with R you can skip ahead**

R is a giant calculator.  There are commands for **mean**, **median**, **standard deviation** etc.  The **summary()** command creates a handy summary, including the mean and median, of all columns in a dataframe.


### Overall summary

Whole dataframe
```{r eval = F}
summary(my.frogs)
```

We can look at just a single column by specifying it using the syntax "dataframe$column.names" where the dataframe and column are sperated by a dollar sig ($).  (note that it prints it out horizontally, not vertically)

```{r}
summary(my.frogs$mass)
```


We used the **make_my_data2L()** command to make a unique subset of the data.   compare the mass values in your subset to the original data

```{r}
summary(my.frogs$mass)
summary(frogarms$mass)
```

**End review section**


----------

### Optional: stacking things with rbind()

**This section is optional**

Handy trick: stack up  summaries with **rbind()**, which stand for "row-bind".

```{r}
rbind(summary(my.frogs$mass),  #note the comma
      summary(frogarms$mass))
```

You can even flip them on their side like this

First, make an object with your summaries
```{r}
my.summaries <- rbind(summary(my.frogs$mass),
                      summary(frogarms$mass))
```

Flip them with the **t()** command ("t" stands for "transpose")
```{r}
t(my.summaries)
```


**End optional section**

----------

### Individual summary stats

**This section is review.  If you are familar with R you can skip ahead**

You can get individual summary statistics using various commands named after the statistic.

The mean of a column with **mean()**.
```{r}
mean(my.frogs$mass)
```


The variance with **var()**.
```{r}
var(my.frogs$mass)
```


Other include: 

* median
* min, max, range
* var, sd
* nrow or length() (for sample size)

Note that **range()** returns 2 values in a **vector**
```{r}
range(my.frogs$mass)
```

### The standard error (SE) in R 

Note that R doesn't return a very common statistic, the **standard error (SE)**. The SE is the standard deviation (SD) divided by the square root of the sample size.  You can get the same size using the length() command.

You can therefore calculate the SE like this: 

```{r}
sd(my.frogs$mass)/sqrt(length(my.frogs$mass))
```


----------

### OPTIONAL: Find a package the calcualtes the SE [O]

**This section is optional**

In the following 2 optional sections you can

* try to find a package with an SE function
* try to write a function that calculates the SE for you


Since R lacks a an SE function many packages include it.  For example, the **plotrix** package has a function **std.error()**.  See if you can download the package, install it using **library()**, and use **std.error()**.  See the help file for more info (?std.error).

Try to look at the underlying code either in the console or by running the debugger using debugonce().

### OPTIONAL: Write your own SD function [O]

Write a function for calculating the SD

Here's a function that takes a single argument "dat_column"
```{r}
#NOTE: this is optional
my_sd1 <- function(dat_column){
  sd(dat_column)/sqrt(length(dat_column))
}
```

To use it, you need to give it the dataframe and the column separated by a "$""

```{r}
my_sd1(my.frogs$mass)
```

Here's a function that takes 2 arguments: the dataframe, and the name of the column  Note that the name of the column needs to be in quotes


```{r}
my_sd2 <- function(dat, column){
  sd(dat[,column])/sqrt(length(dat[,column]))
}

```

You can use the function like this:
```{r}
my_sd2(my.frogs, "mass") #note the use of quotes "..."
```


Here's a fancier function that let's you specify how much to round off the results.  I've set the default rounding to 3 digits.

```{r}
my_sd3 <- function(dat, column, digits.round = 3){
  se <- sd(dat[,column])/sqrt(length(dat[,column]))
  round(se, digits = digits.round)
}

```

The function runs like this.
```{r}
my_sd3(my.frogs, "mass")
```


Note that in all of functions as long as I give the function the **arguments** in the same order they are set up in the code that defines the function, I don't need to provide the agruement names.  This save typing.  Compare these results

```{r eval=FALSE, echo=T}
my_sd3(my.frogs, "mass")
my_sd3(dat = my.frogs, column =  "mass")
my_sd3(column =  "mass", dat = my.frogs)

```

Now try this
```{r make.my_sd3.fail, eval = F}
my_sd3("mass", my.frogs)
```


Can you figure out what has happend with the last one?

**End optional section**

----------

  
## A 1st encounter with dplyr [_]
  
**dplyr** is a package that provides numerous functions for manipulating data.  It is part of the expanding **tidyverse** of packages sponsored in large part by RStudio.  Hadley Whickham is the primary achitect of the tidyverse; he wrote many of the first packages in this framework and laidout the overall conceptual basis that other package authors follow.

For more on dplyr see

* https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html
* https://dplyr.tidyverse.org/
* http://genomicsclass.github.io/book/pages/dplyr_tutorial.html

We will use 2 **dplyr** handy functions

* summarize() / summarise()
* group_by()

**dplyr** can use a syntax that involves "**pipes**".  This is a relatively recent innovation in R coding.  You can string together R commands using the **pipe function**, **%>%**.

Note that the pipe function actually is implemented by the **magrittr** package. If you haven't loaded ggplot, dplyr, or wildlifeR yet you might have to load up magrittr directly.
```{r eval = F}
library(magrittr) 
```

For more background info on pipes see

* http://r4ds.had.co.nz/pipes.html
* https://seananderson.ca/2014/09/13/dplyr-intro/
  
When using **pipes** from **magrittr**, you start with data and follow it with an action you want done to it.  So, for example, *previously* when we wanted the mean of the "mass"" column we did this

```{r eval = F}
mean(my.frogs$mass) #[_]
```

Which is kind of read like a normal mathematical equation or function, where you start from inside the parentheses and work out.  

Eg, this.is.read.2nd(this.is.read.1st)

R let's you nest as many functions as you want.  If I want to round off my calculation I can wrap "mean(my.frogs$mass)" in "round(...)""

```{r}
round(mean(my.frogs$mass)) #[_]
```

Using **pipes** to get the mean I write things more like a sentence

Eg, this.is.read.1st %>% this.is.read.2nd

```{r}
my.frogs$mass %>% mean #[_] note parentheses after mean!
```

Which reads kind of like "Take the mass column and the dataframe and apply the **mean()** function to it."    

To round the mean we would do this

```{r}
my.frogs$mass %>% mean %>% round #[_]
```

Which read left to right like a sentence is "Take the mass column, calculate the mean, and then round off the mean"

Note that the round() command has an argument for how many digits you want to round to.  You include that in the parentheses

```{r}
my.frogs$mass %>% mean() %>% round(digits = 2) #[_]
```


------------

### Optional: Piping everything [O]

**This section is optional**

Most people learn about pipes when doing data summarizing and cleaning with dplyr and friends.  But pipes can be used in many (most?) context. 

Try this

```{r eval = F}
my.frogs$mass %>% hist
```

Not everythign works though.  For example, I can't figure out how to use pipes and t.test().  THere might be a way.
```{r eval = F}
my.frogs %>% t.test(mass ~ sex)
```


**End optional section**

-----------

#### dplyr's summarize() command [_]

Instead of mean(data$column) we can use **summarise()** (for the British) or **summarize()**, plus pipes.

We can get the grand mean of just the mass column by loading **dplyr** using **library()** and then using the **summarise()** command
```{r}
library(dplyr)                      #[_]
my.frogs %>% summarise(mean(mass))
```

This is maybe more complicated than "mean(my.frogs$mass)" or my.frogs$mass %>% mean,  but overall the pipe framework and summarise pays off  when combined with group_by() in the next section

## dplyr's group_by() function [_]

For some more info on **group_by()** see

* https://www.r-bloggers.com/using-r-quickly-calculating-summary-statistics-with-dplyr/
* https://www3.nd.edu/~steve/computing_with_data/24_dplyr/dplyr.html
http://www.datacarpentry.org/R-genomics/04-dplyr.html

We can use group_by() to split things up by a **categorical variable** (sex, color, year).  Here, we can say "take my.frogs, split up the data by the sex column, and apply the mean() function to each subset."  

```{r}
my.frogs %>%            #[_] the data
  group_by(sex) %>%     #the group_by() function applied to the sex column
  summarise(mean(mass)) #the mean() function, applied to mass.
```

This might be a bit abstract when you first do it.  Again, where starting with our whole dataframe (my.frogs), then its piped with %>% over to group_by() function, which splits it essentially into a male and a female sub-dataset.  Then these two subsets are piped again to mean().  Then the mean() function is applied to the mass column in each of these subsets. 

Note that the column heading in the output `mean(mass)`, which is what is in summarise().

Also note that the output is a "tibble", which is a common feature of the tidyverse.  One thing that tibbles do is some reasonable rounding for you automatically.

A handy thing about summarise() is you can pass it labels.  The following code adds a sensible label by changing "summarise(mean(mass))" to "summarise(mass.mean = mean(mass))", where "mass.mean = ..." defines the label.

```{r}
my.frogs %>%                         #[_]
  group_by(sex) %>%
  summarise(mass.mean = mean(mass))
```

You can label things anything, eg "puppies".
```{r}
my.frogs %>%                      #[_]
  group_by(sex) %>%
  summarise(puppies = mean(mass))
```

When I first started using dplyr I found this syntax confusing because people often use the name of the function for the name of the column heading, like this

```{r eval = F}
my.frogs %>%                    #[_]
  group_by(sex) %>%
  summarise(mean = mean(mass))
```


For some reason this trips me up because the word "mean" as a label here is not quoted; to me the word mean should be reserved for the function mean().  Also, in general you should always make both input and output output self-lableing.  If you just look at the output, its not obvious what the "mean" being shown is.


You can pass any summary function to summarise().  We can give it **sd()** to get the sd of mass by sex.  Note that I define the column names using "mass.sd = ..."

```{r}
my.frogs %>%                     #[_]
  group_by(sex) %>%
  summarise(mass.sd = sd(mass))
```


What makes dplyr::group_by and summarize() really powerful is that you can pass it *multiple*. summary functions at the same time.  Here, I'll pass mean() and sd(), naming both.
```{r}
my.frogs %>%                        #[_]
  group_by(sex) %>%
  summarise(mass.mean = mean(mass), #give me the mean!
            mass.sd = sd(mass))     #give me the sd!
```


dplyr also has a handy function **n()** for getting your sample size.
```{r}
my.frogs %>%                       #[_]
  group_by(sex) %>%
  summarise(mass.mean = mean(mass),
            mass.sd = sd(mass),
            n = n())
```


--------------

### OPTIONAL: Using novel functions with dplyr [O]

**This section is optional**

If you have defined the my_sd1() function above you can pass it to summarise() too.

```{r}
my.frogs %>% 
  group_by(sex) %>%
  summarise(mass.mean =  my_sd1(mass))

```


**End optional section**

--------------
  
  
  
------------

## OPTIONAL: Alternatives to dplyr 

**This section is optional**

Most everybody is switching to dplyr.  Below are some other idioms you may see others use or encounter in older books (or my old code).  
  
### doBy::summaryBy

The **doBy** package has a nice syntax.  I don't really see many people use it.  Be sure to download it first.

```{r eval = FALSE}
library(doBy)
summaryBy(mass ~ sex,data = my.frogs, FUN = mean)

summaryBy(mass ~ sex,data = my.frogs, FUN = c(mean,sd))
```


### tapply()

tapply is pretty old school.  You might see people working with big dat get into arguements about whether its faster than dplyr.  
```{r eval = F}
tapply(X = my.frogs$mass,INDEX = my.frogs$sex, FUN = mean)
```


### reshape2::dcast

What I've used most of my career thus far.  Am slowly switch to dplyr.

```{r , last.chnk.sxn4.ch1, eval = F }
library(reshape2)
dcast(data = my.frogs,
      formula = sex ~ .,
      value.var = "mass",
      fun.aggregate  = mean)
```


**End optional section**

------------
  
  