---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Data analysis case study part II: plotting your data


```{r , include=F, echo = F}
knitr::opts_chunk$set(tidy = F, 
                      warning = F,
                      error = F,
                      message = F,
                      #cache = TRUE,
                      fig.margin = TRUE,
                      results="hide",
                      fig.width = 4,
                      fig.height = 3)
```

| **Nathan Brouwer, Phd**
| brouwern@gmail.com
| https://github.com/brouwern
| Twitter: lobrowR

## Introduction

In the previous lesson we made boxplot to explore the distribution of the data.  In this lessons we'll focus on plotting the mean with error bars.  Different kinds of error bars are possible, including the standard deviation,  standard error, and confidence intervals.  Norms for what to plot vary between fields; generally speaking standard error are most common in biology, but 95% CIs are what are recommended by statisticians.  As we'll explain, standard deviations convey one time of information, while standard errors and confidence intervals convey a very different kind.

### Goals & objectives

Create plots to explore the wildlifeR::frogarms data and visualize both variation within the groups using the standard deviation, and how precisely the mean can be estimated using standard errors (SE) and confidence intervals.


### Packages 

* ggplot2
* cowplot
* ggpubr
* dplyr


### Outline

* Preliminaries (if needed): load packages, load data, subset data 
* Represent variation with the standard deviation
* Represent precision of estimated means using SE and 95% CIs


### Vocab

* standard deviation
* standard error
* confidence intervals


### Preliminaries

>The following code doesn't need to be run if the previous exercises in this section were alread run.

#### Load packages

If not already loaded, we need the wildlifeR package, which lives on Github
```{r eval = F}
library(devtools)
install_github("brouwern/wildlifeR")

library(wildlifeR)
```

We also need several other packages needed for visualization.  We'll use the ggplot2 package for plotting, and the cowplot package for some nice plotting defaults.  
```{r}
library(ggplot2)
library(cowplot)
library(ggpubr)
library(dplyr)
```

#### Load data

Load the frogarms data by Buzatto et al (2015) if its not already loaded.
```{r}
data(frogarms)
```


#### Subset your data

The function make_my_data2L() will extact out a random subset of the data.  Change "my.code" to your school email address, minus the "@pitt.edu" or whatever your affiliation is.  **This does not need to be done if you did this as part of the previous lesson**

```{r }
my.frogs <- make_my_data2L(dat = frogarms, 
                           my.code = "nlb24", # <=  change this!
                           cat.var = "sex",
                           n.sample = 20, 
                           with.rep = FALSE)

```


## Background: measures of variation

### Variance

The **variance** is a ubiqitous metric which quantifies the amount of variability in a given set of data.  The equation in all its glory is:

\[s^2 = {\frac{\sum\limits_{i=1}^{n} \left(Y_{i} - \bar{Y}\right)^{2}} {n-1}}\] 

This can be disorienting if you aren't used to looking at this notation.  We'll go over it for the sake of being thorough; we mostly need a general conceptual understanding of the variance right now.

$Y_{i}$ represents each data point, where "i" stands for "index".  $Y_{1}$ is the 1st value in the dataset, $Y_{2}$ is the second values in the data set etc.   The Y with the bar over it, $\bar{Y}$ , is called "**Y bar**" and can be typed out as "Y.bar" in R.  **n** is the sample size. $\sum$ is the summation operator (the Greek letter sigma).  $\left(Y_{i} - \bar{Y}\right)$ is the difference between each value in the data (the Y.i) minus the mean value (the Y.bar).  

$(Y_{i} - \bar{Y})^{2}$ indicates that each difference between a value and the mean should be squared.  These are called the squared differences or squared deviations.  $\sum(Y_{i} - \bar{Y})^{2}$ indicates that all of these values should be summed together; this is called "**the sum of squares**".  Finally, the sum of is divided by n-1.

So, the upshot is: we take each value in the dataset, subtract the mean from each value, square each of those differences, add them up, and divide them by the sample size minus 1.

The variance shows up everywhere in stats, but mostly as part of calculations; less frequently is it reported, though it plays an important role in, among other things, genetics and stochastic demography.  The bigger the variance, the more variability.  However, its "dimensionless." That is, it doesn't have units like meters, pounds, etc.  This is because of the squaring that occurs in the numerator.  The variance is therefore never plotted alongside the datae; this would be meaningless.

The variance of the frogarm mass data is
```{r}
var(frogarms$mass)
```



### Standard deviation (SD)

The variance is a bit hard to wrap you mind around; easier and much more frequently reported in practice is the **standard deviation**.  The standard deviation is just the square root of the variance.

\[\sigma = \sqrt{\frac{\sum\limits_{i=1}^{n} \left(Y_{i} - \bar{Y}\right)^{2}} {n-1}}\]

Taking the square root of the variance puts it back into the same scale as the raw data.  The standard deviation can therefore be plotted alongside the raw data or the mean.

THe standard deviation for the frogarm mass data is

```{r echo = F}
sqrt(var(frogarms$mass))
```

or equivalently
```{r}
sd(frogarms$mass)
```


Since the standard deviation is the square root of the variance, it is always smaller than the variance (Eg s < s^2)

The standard error (and the variance) are both measures of **variation**.  They indicate the amount of variation that occurs in the raw data.  

-------------

### Optional: Testing the equivalence of two values

**THe following is optional**

We can confirm that two values are equivalent like this:

```{r}
sqrt(var(frogarms$mass)) == sd(frogarms$mass)
```


**End optional section**

--------------



### Standard error (SE)

The standard error is often neglected in elementary stats work but it is a central concept.  The standard deviation is the standard deviation divided by the square root of the same size.

\[SE = \frac{SD}{\sqrt{N}}\]

A very important thing about the standard error: it is **not** a measure of variation.  It does **not** indicate how much variation there is in the raw data.  The SE is an indication of **precision**.  Specifically, it indicates how confident we are about our estimate of mean from the data.  The SE does depend on the amount of variation in the data, but also depends on the sample size (n).

Base R does not have a function for the SD.  We can calculate it like this

```{r}
sd(frogarms$mass)/sqrt(length(frogarms$mass))
```

THis is a bit complicated, so we can break it up.  First, assign the numerator and denominator to seperate **R objects**.
```{r}
frog.sd <- sd(frogarms$mass)
frog.n <- sqrt(length(frogarms$mass))
```

Then do the division
```{r echo = F}
frog.sd/frog.n
```

The standard error is frequently used as error bars for plots of means.

### Confidence Interval (CI)

Confidence intervals are a deep topic in stats.  We'll just toch on them briefly.  If you have calcualted the standard error (SE) you can approximate the 95% CI for a mean as

\[CI = 1.96*SE\]

That is, 1.96 (sometimes rounded just to 2.0) multipled by the SE.  The 95% is argueably a better choice for an error bar around a mean than that SE (better yet plot both, which down the row we'll show how to do)


## Representing variation with the SD

ggpubr has a hand function for calcualting means and plotting error bars around them.  The ggerrorplot() is the main function, and the "desc_stat = ..." arguement defines what exactly to plot.

If for some reason you just want to plot means use "desc_stat = 'mean'", with mean quoted.
```{r echo = F}
ggerrorplot(data = my.frogs,
            desc_stat = "mean",
          y = "mass",
          x = "sex")
```

For the standard deviation use "desc_stat = 'mean_sd'" (NOte that any time you have a plot with errorbarrs -- and if you plot means you should have error bars -- you need to define at least in the figure legend what the error bars are.)
```{r}
ggerrorplot(data = my.frogs,
            desc_stat = "mean_sd",
          y = "mass",
          x = "sex")
```

Again, the standard deviation is a measure of **variation**.




### Representing precision with the SE

ggerrorplot() actually defaults to making a plot of the mean +/- 1 standard error.
```{r}
ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex")
```



Mean and 95% confidence interval are plotted with "desc_stat = 'mean_ci' "
```{r}
ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex",
          desc_stat = "mean_ci")
```

Again, the SE and 95%  CI are measure of **precision**.  There can be lots of variation in a dataset (SD is high) but if you have collected a lot of data, you should be able to estimate the mean with precision.  In general, the more data you have, the more precise your estimate will be.  

You can see how more data increases precision by comparing the entire frogarms dataset against your personal subset.  We can assign each plot to an object using the assignment operator "<-" and then plot them side by side with plot_grid() from cowplot.

```{r}
#your data
gg.my.frogs <- ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex",
          desc_stat = "mean_ci",
          ylim = c(1.25,3.75))

#all of the data
gg.all.frogs <- ggerrorplot(data = frogarms, #change data
          y = "mass",
          x = "sex",
          desc_stat = "mean_ci",
          ylim = c(1.25,3.75))

```


Now plot them together.  Then means will be different because of random variation between the subsamples.  What happens to the error bars?.
```{r}
plot_grid(gg.my.frogs, 
          gg.all.frogs,
          labels = c("my frogs","all frogs"))
```


Now, what if instead of the 95% CI we plotted the SD.  What do you think that will look like?  Adapt the code from above by changing " desc_stat = "mean_ci"" to " desc_stat = "mean_sd"".  



## Refining plots

Set colors
```{r}
ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex",
          desc_stat = "mean_ci",
          color = "sex")
```


Add raw data.  kinda crazy but worth seeing how it looks.
```{r}
ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex",
          desc_stat = "mean_ci",
          color = "sex",
          shape = "sex",
          add = "point")
```


Jitter raw data.  Even crazier.  In general it works best if there are less than 10 data points per group.
```{r}
ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex",
          desc_stat = "mean_ci",
          color = "sex",
          add = "jitter")
```




Change to the SD.  In theory, about 2/3 of the data points should falle within +/- 1 SD.  Does that look about right?
```{r }
ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex",
          desc_stat = "mean_sd",
          color = "sex",
          add = "jitter")
```



Back to just the means, and increase size.
```{r}
ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex",
          desc_stat = "mean_ci",
          color = "sex",
          size = 1.5) #
```



Move legend to the bottom using "legend = "bottom" "

```{r}
ggerrorplot(data = my.frogs,
          y = "mass",
          x = "sex",
          desc_stat = "mean_ci",
          color = "sex",
          size = 1.5,
          xlab = "Sex",
          ylab = "Mass (g)",
          legend = "bottom") 
```




