
# (PART) Section II: Plotting data in R with ggplot2 & friends {-}

## {-}

In this section we'll focus on basic plotting skills.  The first chapter in this section is a review of the skills needed to get some data into R from a package.  Subsequent chapters will develop plotting skills with a focus on using ggplot2.  

We'll briefly go over all aspects of **ggplot2** and its varients.  This will orient you to the different things you are likely to see other R users use.  For most of this book, however, we'll focus on a package which creates a "wrapper" for ggplot2 called [ggpubr](http://www.sthda.com/english/rpkgs/ggpubr/) which speeds up the process for beginners of making publication-ready graphs.  Even if you are a ggplot2 pro you should know about ggpubr for making quick graphs, and for teaching others about ggplot2.



# Review: Loading & Examining Data in R

```{r , include=F, echo = F}
knitr::opts_chunk$set(tidy = F, 
                      warning = F,
                      error = F,
                      message = F,
                      cache = F,
                      fig.margin = TRUE,
                      results="hide",
                      fig.width = 3,
                      fig.height = 2)
```

| **Nathan Brouwer, Phd**
| brouwern@gmail.com
| https://github.com/brouwern
| Twitter: lobrowR

## Introduction

This exercise reviews the basics of loading data from a package into R.  These basic skills are needed for the further tasks of plotting that will be built upon in this section.  If you are familiar with R you can probably skim this or skip it altogether.  If you are brand new then this is an excellent review of the material covered so far in this book.

### Learning Goals & Outcomes

To review the basics of loading packages and data.
  
### Functions & Arguements

* data
* library
* ls
* dim
* names
* head
* tail
* is
* summary
* install.packages


### Packages 

* MASS
* cowplot

### Outline

* Intro
* Fisher's Iris data
* Loading data: iris
  + The easy way: from base R w/ data(iris)
* Loading packages in base R: MASS
  + command: library(MASS)
* Loading packages from CRAN: cowplot


## Example data for plotting: Fisher's Irises

* Dataset made popular by R.A. Fisher
* Frequently used to explain/test stats procedures
* See ?iris for more details
* See also https://en.wikipedia.org/wiki/Iris_flower_data_set for more info.


## Loading data into R the easy way: pre-made data in an R "Package"

* Getting data into R (or SAS, or ArcGIS...) can be a pain!
* R comes with many datasets that are pre-loaded into it
* There are also many stat. techniques that can easily be added to R
* These are contained in "packages"

### Load data that is already in the "base" distribution of R

Fisher's iris data comes automatically with R.  You can load it into R's memory using the command "data()"

```{r}
data(iris) #Load the iris data
```

### Look at the iris data

We'll look at the iris data using some commands like ls(), dim(), and names().

You can check that it was loaded using the **ls()** command ("list"). 
```{r, results = "hide"}
ls()
```

You can get info about the nature of the dataframe using commands like **dim()**
```{r, results = "hide"}
dim(iris)
```

This tells us that the iris data is essentially a spreadsheet that has 150 rows and 5 columns.

We can get the column names with names()
```{r}
names(iris)
```

* Note that the first letter of each word is capitalized.  
* What are the implications of this?  

The top of the data and the bottom of the data can be checked with head() and tail() commands

```{r, results = "hide"}
head(iris) #top of dataframe

tail(iris) #bottom of dataframe
```

Another common R command is **is()**, which tells you what something is in R land.
```{r}
is(iris)
```

* R might spew a lot of things out at you when you use **is()**
* usually the 1st item is most important.  
* Here, it tells us that the "object" called "iris" in your workspace is 1st and foremost a "data.frame"
* A dataframe is essentially a spreadsheet of data loaded into R.

You can get basic info about the data themselves using commands like **summary()**.

```{r,results = "hide"}
summary(iris)
```

* Which variables are numeric?
* Which variables are categories/groups (aka "factors")?

If you wanted info on just 1 column, you would tell R to isolate that column like this, using a dollar sign ($).

```{r}
summary(iris$Sepal.Width)
```

That is, that name of the dataframe, a dollar sign ($), and the name of the column.

What happens when you don't capitalize something?  Try these intentional  mistakes (but remove the "#" from in front of each one):

```{r, eval = F}
#all lower case
summary(iris$sepal.width) # this won't work

#just "s" in "sepal" lower case
summary(iris$sepal.Width)  #this won't work either

#or what if you capitalize "i" in "Iris"?
summary(Iris$Sepal.Width) #won't work either
```


The first two error messages are not very informative; the 3rd one ("Error in summary(Iris$Sepal.Width) : object 'Iris' not found") does make a little sense.


## Load data that is in another R package 

### Packages that come with R

* Many scientists develop software for R, and they often include datasets to demonstrate how the software works.  
* Some of this software, called a "package" comes with R already and just needs to be loaded.  
* This is done with the **library()** command.
* The **MASS** package comes with R when you download it and has many useful functions and interesting datasets.

```{r}
library(MASS) #Load the MASS package
```


MASS contains a dataset called called "mammals"
```{r, results = "hide"}
data(mammals)
```


You can confirm that the mammals data is in your workspace using **ls()**
```{r}
ls()
```

You should now have both the "iris"" and the "mammals"" data in your R "workspace.""


What is in the mammals dataset?  Datasets actually usually have useful help files.  Access help using the **?** function.
```{r,results = "hide",echo = F, eval = F}
?mammals
```

The help screen will pop up either within RStudio, or possibly in your web browser.  It tells us that mammals is 

>"A data frame with average brain and body weights for 62 species of land mammals."  

Since this is someone else's data, the authors of the MASS package need to provide proper citation.  At the bottom we can see that these data come from the paper:

>Allison, T. and Cicchetti, D. V. (1976) Sleep in mammals: ecological and constitutional correlates. Science 194: 732-734.


We can learn about the mammals data using the usual commands
```{r eval = FALSE}
dim(mammals)
names(mammals)
head(mammals)
tail(mammals)
summary(mammals)
```

## Load Data From A package On CRAN

Most packages don't come with R when you download it but are stored in a central site called CRAN.  We'll load data from the **cowplot** package.

### Loading packages using R-Studio

RStudio makes it easy to find and load packages.  Follow these instructions.  

* In the panel of RStudio that has the tabs "Plots", "Packages","Help", "Viewer" click on "Packages""
* On the next line it says "Install" and "Update".  Click on "Install"
* A window will pop up.  In the white field in the middle of the window under "Packages" type the name of the package you want.
* RStudio will automatically bring up potential packages as you type.
* Finish typing "cowplot"  or click on the name.
* Click on the "Install" button.
* In the source viewer some misc. test should show up.  Most of the time this works.  If it doesn't, talk to the professor!


If an R packages doesn't load properly, it could be for several reasons. 

1. Your internet connection might be having problems.  
1. The website where the package is stored might be down for maintenance.  
1. The version of are you are using is probably newer than the version of R used to make the package.  This is a real pain - ask for help from an expert R user if think you have this problem.


## Loading packages directly using code

You can also use the **install.packages()** command to try to load the package. 

```{r eval=F}
install.packages("cowplot")
```

## Troubleshooting Package Downloads

### What if you tell R to install a package you already have downloaded?

If you already have the package downloaded to your computer then a window will pop up asking you if you want to restart your computer.  Normally this isn't necessary; just click "no".  You might see a "warning" message pop up in the console such as "Warning in install.packages: package 'cowplot' is in use and will not be installed". This isn't a problem for basic R work.  If you are doing serious work (e.g. for a publication) you should restart R.


### What if I can't get a package I need loaded?

* Talk to someone who is good w/R (eg, your professor)
* Google something like "how to install R package" for general info
* Google something like "problem loading R package"
* Copy and paste any error message you might be getting into Google and see if anyone has written about this problem

See above for reasons why a package might not load properly the 1st time you try.


### Finding R help with Google

There's lots of info about R on the web, and if you have a problem, then someone else has probably had it before and perhaps written something about it.


The website **stackoverflow.com** has lots of info about R.  However, many people who use it are hard-core programmers, who can come across as jerks sometimes when they answer questions if you don't follow the rules and protocols of stackoverflow.

