# Scatterplots in R using qplot()

```{r , include=F, echo = F}
knitr::opts_chunk$set(tidy = F, 
                      warning = F,
                      error = F,
                      message = F,
                      fig.margin = TRUE,
                      results="hide",
                      fig.width = 3,
                      fig.height = 2)
```


## Scatterplots: 2 continuous variables

In this lab we'll explore how to make scatterplots using the qplot() function in ggplot2.

<br>

### R Preliminaries

* We'll use the qplot() function in the *ggplot* package
* The *cowplot* package provides nice deafults for ggplot IMHO

```{r}
library(ggplot2)
library(cowplot)
```


### Scatterplot of Iris data

* Let's make a scatter plot, where we plot 2 continous, numeric variables against each other 
* that is, both x and y variables are numbers; not categories

I've forgotten the names of all the iris variables, so I'll use the **names()** command to see what they are

```{r, results="hide"}
names(iris)
```

<br>

I'll plot the sepals against the petals
```{r, fig.cap = "Sepals vs. Petals"}
qplot(y = Sepal.Length,
      x = Petal.Length, 
      data = iris)
```



### Scatter plot of mammal brain data 

Let's look at another dataset

#### Preliminaries

Get the data from the ggplot2 package

```{r, results="hide"}
data(msleep)
```

#### Look at the data
```{r, results="hide"}
#How much data is there?
dim(msleep)

#What does the data look like
head(msleep)

#Summary of the data
summary(msleep)

```


There are a number of "categorical" varibles in this dataset

* genus 
* vore = carnivore, omnivore et
* order = taxonomic order
* conservation = conservation status (endangered, etc)

For some reason they don't load as "factor" variables (better known as categorical or grouping variables, but called "Factors" in R-land)

We can make these factors using the factor() command

```{r, results="hide"}
msleep$vore <- factor(msleep$vore)
```


Now see what happens when you call summary()
```{r, results="hide"}
summary(msleep)
```


Do the same for "order""

```{r, results="hide"}
msleep$order <- factor(msleep$order)

summary(msleep)
```


And "conservation"

```{r, results="hide"}
msleep$conservation <- factor(msleep$conservation)

summary(msleep)
```

### Make a basic scatterplot
```{r, fig.cap = "Mammal sleep, raw data"}
qplot(y = sleep_total,
      x = brainwt, 
      data = msleep)
```

That looks really really ugly.  It will work better if we "log transform the axes"

<br>

```{r, fig.cap = "Mammal sleep, logged data"}
qplot(y = log(sleep_rem),
      x = log(brainwt), 
      data = msleep)
```

Things get logged all the time in stats.  We'll talk more about that later.


<br>

### Add color coding to scatterplot

```{r, fig.cap = "Add colors with color = "}
qplot(y = log(sleep_rem),
      x = log(brainwt), 
      data = msleep,
      color = vore)
```



### Add color & shape coding to scatterplot

```{r, fig.cap = "Add shapes with shape = "}
qplot(y = log(sleep_rem),
      x = log(brainwt), 
      data = msleep,
      color = vore,
      shape = vore)
```



### Put diffetrent "vores" in seperate panels

* Seperate panels can be made using the "facet" arguement withing qplot

```{r, fig.cap = "Split into different panels w/ facets = "}
qplot(y = log(sleep_rem),
      x = log(brainwt), 
      data = msleep,
      color = vore,
      shape = vore,
      facets = vore ~ .)
```


### Add a "trend line"" to a scatterplot

* Add the geom_smooth() function after the initial qplot() command
* This works best if we remove the "color = vore" command, but you can see what happens if you leave it
```{r, fig.cap = "Add trendline with + geom_smooth()"}
qplot(y = log(sleep_rem),
      x = log(brainwt), 
      data = msleep) +
  geom_smooth()
```


### Exercise: Modify mammal brain code to do the following things

* Change the axes labels
* Add a title
* Use names(msleep) to see what other varibles are in the dataset
* Use summary(msleep) to whether they are continous or categorical
* Pick another continous variable and plot it instead of sleep_total
* Try this with and without logging


