# Scatterplots in R Using qplot()

```{r, first.chnk.sxn3.ch3 , include=F, echo = F}
knitr::opts_chunk$set(tidy = F, 
                      warning = F,
                      error = F,
                      message = F,
                      fig.margin = TRUE,
                      results="hide",
                      fig.width = 3,
                      fig.height = 2)
```

| **Nathan Brouwer, Phd**
| brouwern at gmail.com
| https://github.com/brouwern
| Twitter: lobrowR

## Introduction

We'll show how to make **scatterplots** using the "quick plot" function (**qplot**) from ggplot2.  ggplot2::qplot and ggpubr both offer simplified plotting using tools from ggplot2's toolkit.  ggpubr implements a different syntax, though, while qplot use more standard ggplot2 idioms and so is a good way to get a sense for the full  power of ggploting.

### Learning goals & outcomes {-}

* Make scatter plots : )

  
### Functions & Arguements {-}

* library
* names
* qplot
* data
* dim
* head
* summary
* factor
* log

### Packages {-}

* ggplot2
* cowplot

### Potential hangups {-}

There are many ways to make plots in R: ggplot, qplot, ggpubr, plot; and I'm leaving a few out.  Moving between them is meant to give you a sense for the different tools so you can recognize them "in the wild" on the internet and in books.  We'll transition to focusing on ggpubr soon.


## Scatterplots: 2 Continuous Variables

In this lab we'll explore how to make scatterplots using the qplot() function in ggplot2.



### R Preliminaries

* We'll use the qplot() function in the *ggplot* package
* The *cowplot* package provides nice deafults for ggplot IMHO

```{r, first.chnk.sxn3.ch1, echo = F}
library(ggplot2)
library(cowplot)
```


### Scatterplot of Iris data

* Let's make a scatter plot, where we plot two continous, numeric variables against each other 
* that is, both x and y variables are numbers; not categories

I've forgotten the names of all the iris variables, so I'll use the **names()** command to see what they are

```{r, results="hide"}
names(iris)
```



I'll plot the sepals against the petals
```{r, fig.cap = "Sepals vs. Petals"}
qplot(y = Sepal.Length,
      x = Petal.Length, 
      data = iris)
```



### Scatter plot of mammal brain data 

Let's look at another dataset

#### Preliminaries

Get the data from the ggplot2 package

```{r, results="hide"}
data(msleep)
```

#### Look at the data

```{r, results="hide"}
dim(msleep) #How much data is there?

head(msleep) #What does the data look like

summary(msleep) #Summary of the data
```


There are a number of "categorical" varibles in this dataset

* genus 
* vore = carnivore, omnivore et
* order = taxonomic order
* conservation = conservation status (endangered, etc)

For some reason they don't load as "factor" variables (better known as categorical or grouping variables, but called "Factors" in R-land)

We can make these factors using the factor() command

```{r, results="hide"}
msleep$vore <- factor(msleep$vore)
```


Now see what happens when you call summary()
```{r, results="hide"}
summary(msleep)
```


Do the same for "order""

```{r, results="hide"}
msleep$order <- factor(msleep$order)

summary(msleep)
```


And "conservation"

```{r, results="hide"}
msleep$conservation <- factor(msleep$conservation)

summary(msleep)
```

### Make a basic scatterplot
```{r, fig.cap = "Mammal sleep, raw data"}
qplot(y = sleep_total,
      x = brainwt, 
      data = msleep)
```

That looks really really ugly.  It will work better if we "log transform the axes"
```{r, fig.cap = "Mammal sleep, logged data"}
qplot(y = log(sleep_rem),
      x = log(brainwt), 
      data = msleep)
```

Things get logged all the time in stats.  We'll talk more about that later.




### Add color coding to scatterplot

```{r, fig.cap = "Add colors with color = "}
qplot(y = log(sleep_rem),
      x = log(brainwt), 
      data = msleep,
      color = vore)
```



### Add color & shape coding to scatterplot

```{r, fig.cap = "Add shapes with shape = "}
qplot(y = log(sleep_rem),
      x = log(brainwt), 
      data = msleep,
      color = vore,
      shape = vore)
```



### Put diffetrent "vores" in seperate panels

* Seperate panels can be made using the "facet" arguement withing qplot

```{r, fig.cap = "Split into different panels w/ facets = "}
qplot(y = log(sleep_rem),
      x = log(brainwt), 
      data = msleep,
      color = vore,
      shape = vore,
      facets = vore ~ .)
```


### Add a "trend line"" to a scatterplot

* Add the geom_smooth() function after the initial qplot() command
* This works best if we remove the "color = vore" command, but you can see what happens if you leave it
```{r, last.chnk.sxn3.ch3, fig.cap = "Add trendline with + geom_smooth()"}
qplot(y = log(sleep_rem),
      x = log(brainwt), 
      data = msleep) +
  geom_smooth()
```


### Challenge: Modify mammal brain code 

Modify the mamal bran code to do the following things

* Change the axes labels (eg "+ ylab('y axis')") 
* Add a title (eg  " + ggtitle('...')")
* Use names(msleep) to see what other varibles are in the dataset
* Use summary(msleep) to whether they are continous or categorical
* Pick another continous variable and plot it instead of sleep_total
* Try this with and without logging using the log() command


