---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Plotting continous data in R with ggplot2

| **Nathan Brouwer, Phd**
| brouwern@gmail.com
| https://github.com/brouwern
| @lobrowR

## Introduction

* We're going to  plot some data using the qplot() command
* We'll need to have 2 packages loaded
  + ggplot2, which has the function qplot()
  + cowplot, which provides some nice defaults
* We'll use the iris dataset that comes with R

### Learning objectives


### Learning goals

  
### Functions & Arguements
  

### Packages 

### Files



### Potential hangups


### Outline



```{r , include=F, echo = F}
knitr::opts_chunk$set(tidy = F, 
                      warning = F,
                      error = F,
                      message = F,
                      cache = TRUE,
                      fig.margin = TRUE,
                      results="hide",
                      fig.width = 3,
                      fig.height = 2)
```



## Introduction to ggplot using qplot


```{r, echo = F}
data(iris)

library(ggplot2)
library(cowplot)
```


## A basic plot in ggplot using qplot()

* Unless told otherwise, qplot plots dots

```{r}
qplot(y = Sepal.Length,
      x = Species,    
        data = iris) 
```


# Box plot with labels

* R will usually generate labels for the x and y axes based on the command.  
* These cab be changed by adding another command after the qplot() command
* Add The command **+ xlab("...")** sets the labels for the x-axis, **+ ylab("...")** for the y axis.
* Text for the labels goes in quotes (ie, "Iris species"). 
* THe use of the "+" is different than for most other R packages
* Forgetting the quotes will cause the code to fail.  
* Note that units (mm) are included for the y axis.  



```{r}
qplot(y = Sepal.Length,
      x = Species,    
        data = iris) +
  xlab("Iris Species") +      
  ylab ("Sepal Length (mm)" ) 

```





## Changing colors in R plots

### Changing colors in R plots part 1

* If we wanted we could change the color of the dots using the  argument **"col = "**.  This code can be used to change the color of most types of plots in R.  
* This doesn't increase the information content of the figure but maybe makes it nicer to look at.

```{r}
qplot(y = Sepal.Length,
      x = Species,    
        data = iris,
      color = "red") +
  xlab("Iris Species") +  
  ylab ("Sepal Length (mm)" )

```



### Changing colors in R plots part 2
```{r}
#dopt w/color changes
qplot(y = Sepal.Length,
      x = Species,    
        data = iris,
      color = Species) +
  xlab("Iris Species") +  
  ylab ("Sepal Length (mm)")
```




## Tweaking plots: changing the point size

Run the code below, Can you see what changed?

```{r}
#dopt w/color changes
qplot(y = Sepal.Length,
      x = Species,    
        data = iris,
      color = Species,
      size = 3) +
  xlab("Iris Species") +  
  ylab ("Sepal Length (mm)")

```



## Boxplot with qplot

### Basic boxplot with qplot

* note use of arguement **"geom = ..."**
```{r}
qplot(y = Sepal.Length,
      x = Species,    
        data = iris,
      geom = "boxplot") 
```



## Basic boxplot with colors

* same as before, using "color = "
```{r}
#dopt w/color changes
qplot(y = Sepal.Length,
      x = Species,    
        data = iris,
      geom = "boxplot",
      color = Species) 
```



## Basic boxplot lables

* now use + xlab() and + ylab()
```{r}
qplot(y = Sepal.Length,
      x = Species,    
        data = iris,
      geom = "boxplot",
      color = Species) +
  xlab("Iris Species") +  
  ylab ("Sepal Length (mm)")
```




## Histograms using qplot

* made with geom = "histogram" arguement
* very very easy to make in R with ggplot
* very very very hard to make in Excel
* You should make them all the time for you data!


### Histograms of iris data

* This code makes a histogram of one of the iris species' Petal.Length. 
* Note that you don't have "y = " or "x = " for a histogram!

```{r}
qplot(Petal.Length,
      data = iris)
```



### Histogram with colors

What does this show?

```{r}
qplot(Petal.Length,
      data = iris,
      fill = Species)
```



### Histogram with axes labels

```{r}
qplot(Petal.Length,
      data = iris,
      fill = Species) +  
  xlab ("Sepal Length (mm)")
  
```


### Multiple histograms: "Facets" 

What does this show?

```{r, fig.width = 4, fig.height = 5}
qplot(Petal.Length,
      data = iris,
      fill = Species,
      facets = Species ~.) 
  
```


 
Add a label to x-axis

```{r, fig.width = 4, fig.height = 5}
qplot(Petal.Length,
      data = iris,
      fill = Species,
      facets = Species ~.) +  
  xlab("Sepal Length (mm)")
  
```




## Modifying histograms: titles with the main = argument

* Titles are good for your own personal use but actually are almost never used in figures published in papers and books. 
* We can add a title like this using the arguement "main = "

```{r, fig.width = 4, fig.height = 5}
qplot(Petal.Length,
      data = iris,
      fill = Species,
      main = "Iris species histograms",
      facets = Species ~.) +  
  xlab ("Sepal Length (mm)")
```
