# Loading Excel spreadsheets into RStudio


re-save as .csv and load

load directly

 In this walk through we first re-save this data in an R-compatible format, a "csv" file, called "Lab1_data_PA_eagles.csv".  
 
 
 


### Prepping data in Excel

#### Save data to your R working directory (WD)

Find the file
https://github.com/brouwern/wildlifeR/tree/master/inst/extdata

Save the file "Lab1_data_PA_eagles.xlsx" to your computers desktop.  Today we will be using this as the "working directory"


#### Re-Save The Excel file as a "csv" file

In Excel, follow these steps

+ "File"
+ "Save As"
+ "Browse"
+ Select the working directory (your desktop)
+ Select "Save as type"
+ Select "CSV (Comma delimited)"
+ Click "Save"

The data is now in a format that can be loaded into R.






## Preparing a file for loading into R

Things work best when your Excel file is "clean" & only has exactly what you want in it.  Any extra, accidental typing can cause problems or make things confusing.  A good practice is to always highlight cells to the right of and below your data, right click & select "Delete".  This will remove any accidental typing that occurred. 
Do this to the cells below your data also.


## Reload data

Reload data; be sure to include the "csv" at the end.  Use this code "eaglesPA <- read.csv(file = "eaglesPA.xlsx")".  NOTE: I changed the name of the file to include "_w_2_states" so that I wouldn't overwrite the original file.  Don't use this code unless you changed the file name to the exact same thing

```{r}
#Use this code, w/o the "#" in front of it
# eaglesPA <- read.csv(file = "eaglesPA.xlsx")

#NOTE: I changed the name of the file to include "_w_2_states" so that I wouldn't overwrite the origina file.  Don't use this code unless you changed the file name to the exact smame thing
#eaglesPA <- read.csv(file = "./data/Lab1_data_PA_eaglesPA_w_2_states.csv")

```


Type ls() to see what is now in your workspace
```{r}
ls()
```



Look at the re-loaded eaglesPA data object
```{r eval = FALSE}
summary(eaglesPA)
dim(eaglesPA)
head(eaglesPA)
tail(eaglesPA)
```





