# (PART) SECTION I: Getting software & data into R {-}

In this we'll discuss two of the most fundamental steps of data analysis: getting your hard-earned data into the @$*%# program you want to do you analysis in.  This is often one of the most frustrating steps for beginners because R, like most data analysis tools that aren't spreadsheets, it seemingly very picky about how it wants to receive data.  Luckily there's a method behind the madness, and also some handy tools in RStudio to help with this.

Most data starts off as a spreadsheet before it enters R.  Loading spreadsheet data into R will be our end goal, but as a run up will step through several easier tasks that will highlight the core principles of getting data into R, in particular loading additional software into R (called packages) and loading the data in those packages into R.  We'll also load data directly from the internet into R.  

The chapters in this section are cover the following:

1. Loading R packages to get new software into R
1. Loading and looking at data from R packages
1. Loading packages from the software repository GitHub
1. Loading data from the internet
1. Loading spreadsheets as .csv files
1. Loading Excel spreadsheets

# Loading packages from CRAN

```{r include=FALSE}
knitr::opts_chunk$set(warnings = FALSE, message = FALSE,fig.height = 3, fig.width = 5)
```


| **Nathan Brouwer, Phd**
| brouwern@gmail.com
| https://github.com/brouwern
| @lobrowR

## Introduction

When you install R you get **base R**, which is the core set of functions, functionality, and some data sets.  Base R however is surrounded by a universe of extensions built by statistician, programmers, academics and businesses that use R for analyses.  A lot of R's functionality is found in these packages, including data sets, special plotting functions, and statistical tools for the analysis of complex data.  Some of these are fairly standard and are downloaded along with base R and just need to be explicitly installed.  Other have to be downloaded from the internet and installed.  Most packages contain data in order to demonstrate what they do; working with data from packages will be covered in a later lesson.

This book relies heavily on an R package I've written called "wildlifeR" (https://brouwern.github.io/wildlifeR/) that contains the datasets used throughout the book, as well as some helpful R functions I've written.

Most R packages you'll use are stored on the CRAN website where you download R (https://cran.r-project.org/).  R and RStudio have functions and tools for downloading and managing packages that we'll briefly introduce in this exercise. 

Another place a package can be stored online is a code repository like GitHub.  the wildlifeR package lives on GitHub and can be downloaded directly from there.  Many packages on CRAN also occur on GitHub, especially if programmers are actively developing, updating, and managing the package.  We'll cover downloading packages from GitHub in the next exercise.

### Learning objectives

This exercise will introduce students to

1. the concept of an **R Package** (aka **library**)
1. how to load R packages using the library() function
1. the R plotting package **ggplot2**
1. cool add-ons to ggplot2, **ggpubr** and **cowplot**

### Learning goals

By the end of this exercise students should be able to 
* locate and download packages from the CRAN website using RStudio
* recognize the R functions used to download and install packages.
* 

### Functions & Arguements

* install.packages
  - dependencies = TRUE
* library

### Packages 

* MASS
* ggplot2
* ggpubr
* cowplot

### Potential hangups

* quoted vs. unquoted text (eg qplot vs. ggpubr syntax)


## Loading packages that come with base R

What you download from CRAN is **base R**.  Many functions are called **base functions** because they are hard-wired into R.  

**OPTIONAL: What functions come with base R?**
> If for some reason you want to see *all* the functions that come with base R, type this into the console and press enter.  (ls stands for "list" and is a function we'll use more later).
```{r eval = FALSE}
ls("package:base")
```

As R has been developed there has also built up a cannon of tried and true packages that are downloaded automatically when you download R, but they aren't brought into R's working memory unless you tell R.

**OPTIONAL**: What packages come with base R?

> If you want to see all of the packages that come with base R, do this.  library() is a function you will use a lot.

```{r eval = FALSE}
.libPaths("") 
library()
```

On package that is part of this cannon is MASS, which stands for Modern Applied Statistics in S.  "S" is the precursor to R, and MASS is the package that accompanies the book of the same name, which is one of the original books on S/R.  (https://www.springer.com/us/book/9780387954578)

### [  ] Loading packages with the library() function

When a function is already downloaded to your computer, you explicitly load it into R's working memory using the library() command.
```{r}
library(MASS)
```

### Preview: loading data from packages

Many packages have data.  We can load data using the data() command.
```{r}
data(crabs)
```

We plot data with the plot() command.
```{r}
plot(FL ~ RW, data = crabs)
```

## Load data from an external R package

Many packages have to be explicitly downloaded and installed in order to use their functions and datasets.  Note that this is a **two-step process**:
1. Download package from internet
1. Explicitly tell R to load it


### [ ] Step 1: Downloading packages with install.packages()
There are a number of ways to download packages.  One of the easiest is to use the function install.packages().  Note that it might be better to call this "download.packages" since after you install it, you also have to load it!

Frequently in this book I will include install.packages(...) at the beginning of a lesson the first time we use a package to make sure the package is downloaded.  Note, however, that if you already have downloaded the package, running install.packages(...) will download a new copy.

We'll download a package used for plotting called ggplot2, which stands for "Grammar of Graphics."

```{r eval = FALSE}
install.packages("ggplot2")
```

Often when you download a package you'll see a fair bit of red text, and sometime other things will pop up.  Usually there's nothing of interest here, but sometimes you need to read things carefully over it for hints about why something didn't work.

**OPTIONAL**:Seeing all of your installed packages
*If for some reason you want to see everything you've downloaded, do this.*

```{r eval = F}
installed.packages()
```


### [ ] Step 2: Explicitly loading a package with library()

The install.packages() functions just saves the package software to R; now you need to tell R "I want to work with the package".  This is done using the library() function. (Its called library because another name for packages is libraries)

```{r}
library(ggplot2)
```

**OPTIONAL**: making a plot with ggplot
*Now we can make a plot with ggplot, like using te qplot() function.  (Note that the syntax is different than what we did above with plot() ).*

```{r}
qplot(y=FL,x= RW, data = crabs)
```


## [ ] Downloading packages using RStudio

RStudio has a point-and-click interface to download packages.  In the pane that says "Files, Plots, Packages, Help, Viewer" click on "Packages".  When the panel shift below "Packages" it will say "Install, Update, Packrat."  Click on "Install."  (There might be a lag during this process as RStudio get info about your packages).  In the pop up widow there will be a middle field "Packages" where you can type the name of your package.  There's an auto-complete feature to help you in case you forget the name.  Then click "install."  Note that in the bottom right corner of the pop up is a checked box next to "Install dependencies."  Leave that checked; more on that later.



## [ ] Packages & their dependencies

R packages frequently use other R packages (which frequently use other R packages...).  When an R package requires another package, its called a **dependency.**  Depending on who and how the package was written up, dependencies might not be an issue or could be a problem.

As noted above when you download packages using RStudio's point and click interface there's a box that should be checked called "Install dependencies."

If you are using install.packages() there is an extra argument "dependencies = TRUE" that elicits the same behavior.  I'll use this to get an add-on for ggplot2 called ggpubr.

```{r eval = FALSE}
install.packages("ggpubr",dependencies = TRUE)
```

We can then install this
```{r}
library(ggpubr)
```

**Optional:**Make a plot with ggpubr/
ggpubr is an add on to ggplot.  (This means that ggpubr has ggplot as a dependency).  Note that the syntax for ggpubr function we use, ggscatter(), has a different syntax (again) than ggplot's qplot() function and base R's plot() function.
```{r}
ggscatter(data = crabs,y = "FL", x = "RW") # use quotes!
```


## Challenge

An another add on to ggplot2 is cowplot, which stands for ["Claus O. Wilke Plot"](https://cran.r-project.org/web/packages/cowplot/vignettes/introduction.html).  Download cowplot from CRAN using either the point-and-click method or **install.packages**, and then load it using **library**. Then run the following R code, which should make the plot below.

Note that "FL" and "RW" are NOT in quotation marks as they are for ggscatter()! 

```{r eval = FALSE}
qplot(data = crabs, y = FL, x = RW) #no quotes!
```

```{r echo = FALSE}
library(cowplot)
qplot(data = crabs, y = FL, x = RW)
```

