# Loading data from the internet

| **Nathan Brouwer, Phd**
| brouwern at gmail.com
| https://github.com/brouwern
| Twitter lobrowR

## Introduction

Its possible to download data directly from the internet, including

* Spreadsheets directly posted online as .csv or .txt
* Spreadsheets contained within a GitHub repository, including a package
* Google Sheets
* and many other formats

In this short exercise we'll download some data that is stored as a raw .csv file within the inner workings of the widlifeR package.  

### Learning Goals & Outcomes

By the end of this lesson students should be able to download basic data sources from the internet using getURL() from the RCUrl package.

  
### Functions & Arguements

* RCurl::getURL()
* scatter.smooth

### Packages 

* RCurl
install.packages()

### Potential hangups

* Bad internet connection
* Firewall problems



## [ ] Downloading a .csv file using getURL()

The package RCurl provides functions for accessing online material.  First we need the package

```{r eval = FALSE}
install.packages(RCurl, dependencies = TRUE) # [ ]
```

As always, once we install a package we need to **really** install it with library(). (You might see some red text as RCurl loads up some of its **dependencies**)
```{r}
library(RCurl) # [ ]
```

We then use the getURL() to prep the info we need for downloading that .csv we want.

| The file we want is "eaglesWV.csv".  It is located at this rather long URL:
| https://raw.githubusercontent.com/brouwern/wildlifeR/master/inst/extdata/eaglesWV.csv

First, we'll use the "<-" assignment operator to store the shortened URL in an R object.  Be sure to put the URL in quotes.
```{r}
eaglesWV.url <- "https://raw.githubusercontent.com/brouwern/wildlifeR/master/inst/extdata/eaglesWV.csv" # [ ]
```

Next we'll use the getURL() function to set things up, storing the info in a new object "eaglesWV.url_2"  (note the "_2" on the end).

First, get the URL from the URL-containing object we just made.
```{r}
eaglesWV.url_2 <- getURL(eaglesWV.url)
```

Now use read.csv() to actually get it.
```{r}
eaglesWV_2 <- read.csv(text = eaglesWV.url_2)
```

We can preview the downloaded dataset using summary() or any other command we want

```{r}
summary(eaglesWV_2)
```



----------------------

## OPTIONAL: Plotting West Virginia Eagle Data

**This section is optional** 

Thankfully, eagles having been increasing exponentially in West Virginia since the 1980s.
```{r}
scatter.smooth(y = eaglesWV_2$WV,x = eaglesWV_2$year)
```

**End optional section** 

----------------------


